Proving Logical Atomicity using Lock Invariants\thanks{Supported by the National Science Foundation via grant \#1811894.}

### 2022-05-16 concurrency/ASL 2022/Verification of Concurrent Binary Search Tree.tex
edited proofs

For many concurrent data structures, the ideal correctness condition is that the data structure behaves the same as a sequential implementation, even when accessed simultaneously by multiple threads. This intuitive condition has been formalized in terms of linearizability (operations appear to take effect in some total order) or atomicity (client threads never observe intermediate states of operations). In separation logic, atomicity can be expressed in the form of \emph{atomic triples}~\cite{tada} of the form $$\langle a.\ P_l\ |\ P_p(a)\rangle\ c\ \langle Q_l\ |\ Q_p(a)\rangle$$ where $P_l$ and $Q_l$ are \emph{private} pre- and postconditions similar to an ordinary Hoare triple, and $P_p$ and $Q_p$ are \emph{public} pre- and postconditions, parameterized by an abstract value $a$ of the shared data structure. This triple says that if $P_l$ is true before a call, and the fact that $P_p$ is true for some value of $a$ is held in shared state (e.g. a global invariant), then $P_p$ will continue to be true for some (possibly different) value of $a$ up until the \emph{linearization point} of $c$, at which point $Q_p$ will become true atomically for the same value $a$ (and $Q_l$ will be true after $c$ ends). \LB{This suggests I should interpret the leftmost occurrence of 'a' as a binder whose scope extends all the way to $Q_p(a)$. So maybe the binding should happen outside the left pair of angle brackets?}\wm{In the original TaDA paper they did this with a fancy $\forall$ outside that I can't figure out how to replicate. In Iris, they use the notation I'm using here, even though the scope looks wrong.} \LB{Also, as decorations 'l' and 'p' are not indices but just markers (they don't range over anything) and both 'private' and 'public' start with the letter 'p', maybe replacing $P_l$, $P_p$, $Q_l$, $Q_p$ with say $P$, $Q$, $R$, $S$, or maybe $P$, $P'$, $Q,$, $Q'$ could help. Indeed, the concrete patters below suggest something like $P$, $Q$, $P'$, $Q'$, to emphasize that the two predicates to the left of $\mid$ have something in common, and so do the two predicates to the right of $\mid$.} The specific value of $a$ (the shared abstract state) can vary as other threads modify the data structure, as long as $P_p(a)$ is always maintained. For instance, the specification $$\langle s.\ \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ s\rangle\ \texttt{push}(v)\ \langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ (v :: s)\rangle$$ expresses the fact that the \texttt{push} operation of a concurrent stack correctly implements the behavior of a sequential push, making no visible changes to whatever stack $s$ is in place until, at the linearization point, the current stack $s$ is replaced atomically with $v :: s$. The $\mathsf{stack}$ itself is treated as a publicly owned resource, and can only be accessed and modified atomically by threads holding the $\mathsf{is\_stack}$ assertion.

\subsection{Two Styles of Lock Specification}
Soon after the introduction of concurrent separation logic~\cite{csl}, both Gotsman et al.~\cite{gotsman} and Hobor et al.~\cite{oraclesematic} extended it with support for C-style storable locks, where a lock is a special memory location designated as coordinating access to some other resource. Their lock specifications were essentially equivalent, and of the form:
$$\{ \ell \islock R \}\ \texttt{acquire}(\ell)\ \{ R * \ell \islock R \} \qquad \{ R * \ell \islock R \}\ \texttt{release}(\ell)\ \{ \ell \islock R \}$$
Each lock $\ell$ is associated with a ``lock invariant'' $R$, an arbitrary separation logic predicate representing the protected resource, which is gained by a thread that acquires the lock and must be restored when the lock is released. At the time, there was no formal treatment of atomic operations in CSL, so actual lock implementations were beyond the scope of verification: these specs were taken as axioms, or proved through direct appeal to data-race-free guarantees of specific memory models. In garbage-collected languages that do not deallocate locks, the assertion $\ell \islock R$ can be freely duplicated and shared among threads accessing $R$; if we want to track ownership and eventually deallocate the lock, we can instead split the assertion into fractional shares such that $\ell \islock_{\pi_1} R * \ell \islock_{\pi_2} R = \ell \islock_{\pi_1 + \pi_2} R$, and then allow deallocation only when we own the full share $\ell \islock_1 R$.

Later, the TaDA logic~\cite{tada}, which codified the notion of logical atomicity, presented an alternative set of lock specifications:
$$\langle b.\ (\mathsf{L}(\ell) \land \neg b) \vee (\mathsf{U}(\ell) \land b) \rangle\ \texttt{acquire}(\ell)\ \langle \mathsf{L}(\ell) \land b \rangle \qquad \langle \mathsf{L}(\ell) \rangle\ \texttt{release}(\ell)\ \langle \mathsf{U}(\ell) \rangle$$
An \texttt{acquire} operation may access the lock repeatedly, finding it in either the locked ($\mathsf{L}$) or unlocked ($\mathsf{U}$) state; if it sees the unlocked state ($b$ holds), then it can atomically move to a locked state and return. A release simply moves atomically from the locked state to the unlocked state. These operations make no mention of a protected resource, but because they are atomic, they can interact with any global invariant (``shared region'' in TaDA's terminology): in particular, in combination with the invariant $(\mathsf{U}(\ell) * R) \vee \mathsf{L}(\ell)$, they can be used to derive the lock-invariant-based specs. It is also worth noting that these lock specifications are not axioms: they are proved for a simple spinlock implementation that uses atomic operations (e.g. compare-and-swap), which are the actual primitive operations in TaDA and the logics that follow it. In this style, tracking ownership of a lock is orthogonal to the lock assertion itself; a mechanism such as cancelable invariants~\cite{rustbelt-relaxed} can be used to both make the lock assertion $\mathsf{U}$/$\mathsf{L}$ publicly available and deallocate it once all threads are finished.

@@

For many concurrent data structures, the ideal correctness condition is that the data structure behaves the same as a sequential implementation, even when accessed simultaneously by multiple threads. This intuitive condition can be formalized as linearizability (operations appear to take effect in some total order) or atomicity (client threads never observe intermediate states of operations). In separation logic, atomicity can be expressed in the form of \emph{atomic triples}~\cite{tada}, written in the form $\forall a.\ \langle P_l\ |\ P_p(a)\rangle\ c\ \langle Q_l\ |\ Q_p(a)\rangle$, where $P_l$ and $Q_l$ are \emph{private} pre- and postconditions similar to an ordinary Hoare triple, and $P_p$ and $Q_p$ are \emph{public} pre- and postconditions, parameterized by an abstract value $a$ of the shared data structure. Intuitively, $P_l$ and $Q_l$ must be true before and after the call, while $P_p$ must be true for some value of $a$ at every point from the beginning of $c$ until some designated linearization point, at which point $Q_p$ becomes true atomically for the same value $a$. For instance, the specification $$\forall s.\ \langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ s\rangle\ \texttt{push}(v)\ \langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ (v :: s)\rangle$$ expresses the fact that the \texttt{push} operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transitioning from some stack $s$ to $v :: s$ at some point during its execution. %maybe something about linking ghost state, which parts are real vs. ghost, accessing invariants around atomic triple
%Maybe say what a reasonable $Q$ is and why such a spec is still useful, specifically w.r.t. the situation where one of the other threads' action following the linearization point removes the element just inserted.

@@

\subsection{Atomicity with atomic locks}
Proving atomic specifications for a data structure using atomic specifications for locks is more immediate. As an example, consider Krishna et al.'s verification of a hand-over-hand locking pattern for search structures~\cite{templates}. The lock specification used in the paper is $$\langle \mathsf{inFP}(n)\ |\ \mathsf{CSS}(r, C) \rangle\ \mathtt{lockNode}\ n\ \langle \mathsf{CSS}(r, C) * \mathsf{N}(n)\rangle$$ where $\mathsf{inFP}(n)$ asserts that the node $n$ is in the data structure, $\mathsf{CSS}$ is the abstract state, and $\mathsf{N}(n)$ is the resources for the node $n$ (including both points-to assertions and ghost state). The abstract state assertion $\mathsf{CSS}$ includes both global ghost state and the composition of per-node state $\mathsf{N}$ for each node that is not currently locked, so that $\mathsf{N}(n)$ can be removed from the abstract state when $n$'s lock is acquired.

@@

We demonstrate our approach to proving logically atomic specifications with invariant-based (type 1) lock specs by verifying a binary search tree (BST) with fine-grained locking in VST, using the approach of section~\ref{atomicity} to prove atomic specifications for the BST operations while using invariant-style specs for our locks (as required by VST's soundness proof).
As described in section~\ref{atomic}, we want the operations of our concurrent BSTs to satisfy atomic triples relating them to sequential operations\LB{As these specs were already given on page 5 we don't need to repeat (maybe put them in a Figure on page 5 and then refer to the figure here?}:
\begin{mathpar}
\langle m.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (m[k\mapsto v])\rangle

\langle m.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ m \land m(k) = v\rangle

\langle m.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep\ p\ |\ \treerep\ (m[k\mapsto \_])\rangle
\end{mathpar}
The details of $\nodeboxrep$ and $\treerep$ may vary by implementation, but we expect that $\nodeboxrep$ contains at least a pointer to the data structure, and $\treerep$ contains some ghost state linking the concrete state of the data structure to the abstract key-value map $m$ implemented by the tree. In this section, we describe the construction of these predicates, and the proofs of the operations, for a BST implementation based on hand-over-hand locking, using lock-invariant-based specs for our locks.

@@

Hand-over-hand locking (also called lock coupling) is a fine-grained locking pattern for traversing a data structure in which we acquire the lock for the next node before releasing the lock for the current node, thus guaranteeing that the link between the two nodes is not removed or rearranged while we traverse it. Hand-over-hand locking is a classic example of a pattern that requires reference to lock invariants inside other lock invariants, dating back to the earliest papers on ``predicates in the heap'' in CSL~\cite{gotsman}. The trick is to assign each node in the data structure a lock invariant that includes the lock assertion for its child nodes; for a linked list, for instance, we might write\LB{label this definition as equation (1)} $R(n) \triangleq n \mapsto (d, n', \ell') * \ell' \islock R(n')$, where $n'$ is the child of $n$ and $\ell'$ is $n'$'s lock. By acquiring $n$'s lock, we learn the lock assertion for $n'$, and can acquire its lock as needed before releasing $n$.

%make this more concise, and maybe merge it into a later section
This invariant becomes slightly more complicated when we consider \emph{ownership} of locks, as is required if we want to eventually free the lock and return its resources. In this case the lock assertion carries a share $\pi$ \LB{$\pi$ a rational(real?) number in the interval\ldots}, where the full share $\pi = 1$ is required to free the lock, while any nonzero share is sufficient to acquire or release it. We can amend our lock invariant to $R(n) \triangleq n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n')$ for some suitable share $\pi$, but then, what happens to the share of $n'$'s %awkward
 lock when we release the lock on $n$? Starting from a state in which we hold $n$'s lock $\ell$, this gives us:
$$\begin{array}{l}
\{ \ell \islock_\pi R(n) * n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n') \}\\
\texttt{acquire}(\ell');\\
\{ \ell \islock_\pi R(n) * n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_\pi R(n'') \}\\
\texttt{release}(\ell);\\
\{ \ell \islock_\pi R(n) * n' \mapsto (d, n'', \ell'') * \ell'' \islock_\pi R(n'') \}
\end{array}$$
\LB{At first, I thought that the $\pi$ in $ \ell'' \islock_\pi R(n'')$ should/could be different from the one in $\ell \islock_\pi R(n)$. But the point is that its the same $\pi$, and $\pi$ is a parameter to entire setup. Maybe that's worth saying?}  the In order to re-establish $\ell$'s lock invariant $R(n)$, we must give up our knowledge of $n'$'s lock $\ell'$ \LB{perhaps better:  \ldots we had to givegive up our knowledge of $n'$'s lock $\ell'$ in the second proof step (the call to release) here.}. This leaves us with two problems: we have no lock assertion for $\ell'$, so we will be unable to release it later, and we still have partial ownership of $\ell$'s lock assertion, without any way to re-collect our share. Both of these problems can be solved by making the lock \emph{recursive}, including a share of the lock assertion in its own lock invariant\footnote{Gotsman et al. instead use a separation assertion $\mathsf{Locked}$ that is always held by the thread that acquires a lock, and is required to free the lock: this is equivalent to an extra share of the lock assertion. \LB{Do you mean some thing like Gotsman uses Locked(l), defined by $Locked(l) = l \mapsto_{\pi_1}R(n)$?}}\LB{i.e by replacing definition (1) by the following (and label the definition, too). And again $pi_1$ and $\pi_2$ are parameters, not chosen new each time.)}:
$$R(n) \triangleq \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n')$$

Share $\pi_1$ of each lock is held by the lock itself, while share $\pi_2$ is held by its parent\LB{wait, locks have parents, too? I though just nodes in the tree}. Then the same sequence of operations gives us:
$$\begin{array}{l}
\{ \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n') \}\\
\texttt{acquire}(\ell');\\
\{ \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n') * \ell' \islock_{\pi_1} R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_{\pi_2} R(n'') \}\\
\texttt{release}(\ell);\\
\{ \ell' \islock_{\pi_1} R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_{\pi_2} R(n'') \}
\end{array}$$
Now, when we acquire $\ell'$, we hold both $\pi_1$ and $\pi_2$ of its lock assertion; when we release $\ell$, we return our share $\pi_1$ of $\ell$ along with share $\pi_2$ of $\ell'$, but retain share $\pi_1$ of $\ell'$, so we can release it later. The mechanism for defining a lock invariant that includes a share of its own lock assertion may vary across separation logics (VST provides a $\mathsf{selflock}$ assertion for exactly this purpose), but as long as such a mechanism exists, this pattern can be used to implement hand-over-hand lock invariants with share accounting, allowing us to traverse a data structure built with this pattern and then reclaim all of its resources once all threads are finished with it.

@@

\subsection{Specification of the hand-over-hand BST}
%rewrite to use the above: this is the global state made up of components, etc.

To verify our tree implementation\LB{I think here would be a good place to show the C structs and function prototypes of insert etc. (but not the function bodies)}, we need to instantiate $\nodeboxrep$ (the resources held by each client thread) and $\treerep$ (the abstract state of the data structure), and show how they interact to access and update the state of the data structure for each call. In a lock-based implementation, $\nodeboxrep$ will be a pointer to the lock of the root node, and acquiring that lock will give the thread access to more of the data structure. As described in section \ref{atomicity}, our approach is to associate a piece of ghost state with each node in the tree, which will be shared between that node's lock and the abstract state in $\treerep$. We also need to make sure that our lock invariants support the hand-over-hand pattern.

\subsubsection{Key Ranges}%can cut this down if need be
When we traverse the BST looking for a key $k$, we will reach either a node containing $k$, or an empty node where $k$ would appear if it was in the tree. To prove correctness of this procedure, we need ghost state that tracks where each key ``should appear'' in the current tree. As observed by Krishna et al.~\cite{krishna2017flow}, this can be done by associating each node with a lower and upper bound on the keys appearing in the subtree rooted at that node. We refer to the product of these bounds as a \emph{range}. The range of a node is inherited from its parent, based on the parent's key: if node $n$ has range $(l, r)$ and key $k$, then its left child has range $(l, k)$ and its right child has range $(k, r)$. Figure~\ref{range_bst} shows an example of a BST with each node labeled with its range, starting with $(-\infty, +\infty)$ at the root and propagated to the empty \emph{leaf} nodes.

@@

\begin{figure}[htb]
\centering
%% \forestset{
%%    lab/.style={
%%         label={[anchor=west, scale=0.75] north east:#1},
%%     },
%% }
\begin{forest}
for tree={
  calign=fixed edge angles,
  calign angle=65,
  circle, draw, minimum size=3ex, inner sep=1pt, l sep=0, l=0,
%%     s sep=3mm, l sep=0mm
        }
    % my label/.style={
        % label={[anchor=south west] north east:#1},
    % },
    [ 20, tikz={\node[right=0pt of .north east, scale=0.75]  {$(-\infty,\infty)$};}
        [15, tikz={\node[left=0pt of .north west, scale=0.75]  {$(-\infty,20)$};}]
        [30, tikz={\node[right=0pt of .north east, scale=0.75]  {$(20,\infty)$};}
            [25, tikz={\node[left=-2pt of .west, scale=0.75]  {$(20,30)$};}]
            [50, tikz={\node[right=0pt of .north east, scale=0.75]  {$(30,\infty)$};}
                [45, tikz={\node[right=0pt of .south east, scale=0.75]  {$(30,50)$};}
                    [40, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,45)$};}
                        [35, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,40)$};}
                            [, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,35)$};}]
                            [, tikz={\node[right=0pt of .north east, scale=0.75]  {$(35,40)$};}]
                        ]
                        [,no edge, draw=none]
                    ]
                    [,no edge, draw=none]
                ]
                [,no edge, draw=none]
            ]
        ]
    ]
\end{forest}
\caption{A BST with each node labeled with its range; we omit the empty children of nodes other than 35}
\label{range_bst}
\end{figure}

>>>>

For many concurrent data structures, the ideal correctness condition is that the data structure behaves the same as a sequential implementation, even when accessed simultaneously by multiple threads. This intuitive condition has been formalized in terms of linearizability (operations appear to take effect in some total order) or atomicity (client threads never observe intermediate states of operations). In separation logic, atomicity can be expressed in the form of \emph{atomic triples}~\cite{tada} of the form \[\langle a.\ P_l\ |\ P_p(a)\rangle\ c\ \langle Q_l\ |\ Q_p(a)\rangle\] where $P_l$ and $Q_l$ are \emph{private} pre- and postconditions similar to an ordinary Hoare triple, and $P_p$ and $Q_p$ are \emph{public} pre- and postconditions, parameterized by an abstract value $a$ of the shared data structure. This triple says that if $P_l$ is true before a call, and the fact that $P_p$ is true for some value of $a$ is held in shared state (e.g. a global invariant), then $P_p$ will continue to be true for some (possibly different) value of $a$ up until the \emph{linearization point} of $c$, at which point $Q_p$ will become true atomically for the same value $a$ (and $Q_l$ will be true after $c$ ends). \LB{This suggests I should interpret the leftmost occurrence of 'a' as a binder whose scope extends all the way to $Q_p(a)$. So maybe the binding should happen outside the left pair of angle brackets?}\wm{In the original TaDA paper they did this with a fancy $\forall$ outside that I can't figure out how to replicate. In Iris, they use the notation I'm using here, even though the scope looks wrong.} \LB{Also, as decorations 'l' and 'p' are not indices but just markers (they don't range over anything) and both 'private' and 'public' start with the letter 'p', maybe replacing $P_l$, $P_p$, $Q_l$, $Q_p$ with say $P$, $Q$, $R$, $S$, or maybe $P$, $P'$, $Q,$, $Q'$ could help. Indeed, the concrete patters below suggest something like $P$, $Q$, $P'$, $Q'$, to emphasize that the two predicates to the left of $\mid$ have something in common, and so do the two predicates to the right of $\mid$.} The specific value of $a$ (the shared abstract state) can vary as other threads modify the data structure, as long as $P_p(a)$ is always maintained. For instance, the specification \[\langle s.\ \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ s\rangle\ \texttt{push}(v)\ \langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ (v :: s)\rangle\] expresses the fact that the \texttt{push} operation of a concurrent stack correctly implements the behavior of a sequential push, making no visible changes to whatever stack $s$ is in place until, at the linearization point, the current stack $s$ is replaced atomically with $v :: s$. The $\mathsf{stack}$ itself is treated as a publicly owned resource, and can only be accessed and modified atomically by threads holding the $\mathsf{is\_stack}$ assertion.

\subsection{Two Styles of Lock Specification}
Soon after the introduction of concurrent separation logic~\cite{csl}, both Gotsman et al.~\cite{gotsman} and Hobor et al.~\cite{oraclesematic} extended it with support for C-style storable locks, where a lock is a special memory location designated as coordinating access to some other resource. Their lock specifications were essentially equivalent, and of the form:
\[\{ \ell \islock R \}\ \texttt{acquire}(\ell)\ \{ R * \ell \islock R \} \qquad \{ R * \ell \islock R \}\ \texttt{release}(\ell)\ \{ \ell \islock R \}\]
Each lock $\ell$ is associated with a ``lock invariant'' $R$, an arbitrary separation logic predicate representing the protected resource, which is gained by a thread that acquires the lock and must be restored when the lock is released. At the time, there was no formal treatment of atomic operations in CSL, so actual lock implementations were beyond the scope of verification: these specs were taken as axioms, or proved through direct appeal to data-race-free guarantees of specific memory models. In garbage-collected languages that do not deallocate locks, the assertion $\ell \islock R$ can be freely duplicated and shared among threads accessing $R$; if we want to track ownership and eventually deallocate the lock, we can instead split the assertion into fractional shares such that $\ell \islock_{\pi_1} R * \ell \islock_{\pi_2} R = \ell \islock_{\pi_1 + \pi_2} R$, and then allow deallocation only when we own the full share $\ell \islock_1 R$.

Later, the TaDA logic~\cite{tada}, which codified the notion of logical atomicity, presented an alternative set of lock specifications:
\[\langle b.\ (\mathsf{L}(\ell) \land \neg b) \vee (\mathsf{U}(\ell) \land b) \rangle\ \texttt{acquire}(\ell)\ \langle \mathsf{L}(\ell) \land b \rangle \qquad \langle \mathsf{L}(\ell) \rangle\ \texttt{release}(\ell)\ \langle \mathsf{U}(\ell) \rangle\]
An \texttt{acquire} operation may access the lock repeatedly, finding it in either the locked ($\mathsf{L}$) or unlocked ($\mathsf{U}$) state; if it sees the unlocked state ($b$ holds), then it can atomically move to a locked state and return. A release simply moves atomically from the locked state to the unlocked state. These operations make no mention of a protected resource, but because they are atomic, they can interact with any global invariant (``shared region'' in TaDA's terminology): in particular, in combination with the invariant $(\mathsf{U}(\ell) * R) \vee \mathsf{L}(\ell)$, they can be used to derive the lock-invariant-based specs. It is also worth noting that these lock specifications are not axioms: they are proved for a simple spinlock implementation that uses atomic operations (e.g. compare-and-swap), which are the actual primitive operations in TaDA and the logics that follow it. In this style, tracking ownership of a lock is orthogonal to the lock assertion itself; a mechanism such as cancelable invariants~\cite{rustbelt-relaxed} can be used to both make the lock assertion $\mathsf{U}$/$\mathsf{L}$ publicly available and deallocate it once all threads are finished.

@@

For many concurrent data structures, the ideal correctness condition is that the data structure behaves the same as a sequential implementation, even when accessed simultaneously by multiple threads. This intuitive condition can be formalized as linearizability (operations appear to take effect in some total order) or atomicity (client threads never observe intermediate states of operations). In separation logic, atomicity can be expressed in the form of \emph{atomic triples}~\cite{tada}, written in the form $\forall a.\ \langle P_l\ |\ P_p(a)\rangle\ c\ \langle Q_l\ |\ Q_p(a)\rangle$, where $P_l$ and $Q_l$ are \emph{private} pre- and postconditions similar to an ordinary Hoare triple, and $P_p$ and $Q_p$ are \emph{public} pre- and postconditions, parameterized by an abstract value $a$ of the shared data structure. Intuitively, $P_l$ and $Q_l$ must be true before and after the call, while $P_p$ must be true for some value of $a$ at every point from the beginning of $c$ until some designated linearization point, at which point $Q_p$ becomes true atomically for the same value $a$. For instance, the specification \[\forall s.\ \langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ s\rangle\ \texttt{push}(v)\ \langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ (v :: s)\rangle\] expresses the fact that the \texttt{push} operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transitioning from some stack $s$ to $v :: s$ at some point during its execution. %maybe something about linking ghost state, which parts are real vs. ghost, accessing invariants around atomic triple
%Maybe say what a reasonable $Q$ is and why such a spec is still useful, specifically w.r.t. the situation where one of the other threads' action following the linearization point removes the element just inserted.

@@

\subsection{Atomicity with atomic locks}
Proving atomic specifications for a data structure using atomic specifications for locks is more immediate. As an example, consider Krishna et al.'s verification of a hand-over-hand locking pattern for search structures~\cite{templates}. The lock specification used in the paper is \[\langle \mathsf{inFP}(n)\ |\ \mathsf{CSS}(r, C) \rangle\ \mathtt{lockNode}\ n\ \langle \mathsf{CSS}(r, C) * \mathsf{N}(n)\rangle\] where $\mathsf{inFP}(n)$ asserts that the node $n$ is in the data structure, $\mathsf{CSS}$ is the abstract state, and $\mathsf{N}(n)$ is the resources for the node $n$ (including both points-to assertions and ghost state). The abstract state assertion $\mathsf{CSS}$ includes both global ghost state and the composition of per-node state $\mathsf{N}$ for each node that is not currently locked, so that $\mathsf{N}(n)$ can be removed from the abstract state when $n$'s lock is acquired.

@@

We demonstrate our approach to proving logically atomic specifications with invariant-based (type 1) lock specs by verifying a binary search tree (BST) with fine-grained locking in VST, using the approach of section~\ref{atomicity} to prove atomic specifications for the BST operations while using invariant-style specs for our locks (as required by VST's soundness proof).
As described in section~\ref{atomic}, we aim to define assertions $\nodeboxrep$ (the per-thread handle to the BST) and $\treerep$ (the shared abstract state) such that the operations of the concurrent BST satisfy atomic triples along the lines of \[\langle m.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (m[k\mapsto v])\rangle\] In this section, we describe the construction of the two predicates and the proofs of the operations for a BST implementation with hand-over-hand locking, using lock-invariant-based specs for our locks.

@@

Hand-over-hand locking (also called lock coupling) is a fine-grained locking pattern for traversing a data structure in which we acquire the lock for the next node before releasing the lock for the current node, thus guaranteeing that the link between the two nodes is not removed or rearranged while we traverse it. Hand-over-hand locking is a classic example of a pattern that requires lock invariants that refer to other lock invariants, dating back to the earliest papers on ``predicates in the heap'' in CSL~\cite{gotsman}. The trick is to assign each node in the data structure a lock invariant that includes the lock assertion for its child nodes; for a linked list, for instance, we might write \begin{equation}\label{inv1}R(n) \triangleq n \mapsto (d, n', \ell') * \ell' \islock R(n')\end{equation} where $n'$ is the child of $n$ and $\ell'$ is $n'$'s lock. By acquiring $n$'s lock, we learn the lock assertion for $n'$, and can acquire its lock as needed before releasing $n$.

%make this more concise, and maybe merge it into a later section
This invariant becomes slightly more complicated when we consider \emph{ownership} of locks, as is required if we want to eventually free the lock and return its resources. In this case the lock assertion carries a share $\pi$ (often represented as a rational number in the range $(0, 1]$), where the full share $\pi = 1$ is required to free the lock while any share is sufficient to acquire or release it. We can amend our lock invariant to \begin{equation}R(n) \triangleq n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n')\end{equation} for some fixed share $\pi$, but then, what happens to the share of $\ell'$ when we release the lock on $n$? Starting from a state in which we hold $n$'s lock $\ell$, this gives us:
\[\begin{array}{l}
\{ \ell \islock_\pi R(n) * n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n') \}\\
\texttt{acquire}(\ell');\\
\{ \ell \islock_\pi R(n) * n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_\pi R(n'') \}\\
\texttt{release}(\ell);\\
\{ \ell \islock_\pi R(n) * n' \mapsto (d, n'', \ell'') * \ell'' \islock_\pi R(n'') \}
\end{array}\]
In the call to \texttt{release}, we give up resources to reestablish $\ell$'s lock invariant, including the lock assertion for $\ell'$. As a result, we are left with a share of $\ell$ (the lock for the node we left behind) and no share of $\ell'$ (the lock for the current node).

Both of these problems can be solved by making the lock \emph{recursive}, including a share of $\ell$ in $\ell$'s own lock invariant\footnote{Gotsman et al. instead use a separation assertion $\mathsf{Locked}$ that is always held by the thread that acquires a lock, and is required to free the lock: this is functionally equivalent to an extra share of the lock assertion. \LB{Do you mean some thing like Gotsman uses Locked(l), defined by $Locked(l) = l \mapsto_{\pi_1}R(n)$?}\wm{It's not defined that way---it's a separate assertion---but it could be implemented that way.}}. We split ownership of the lock $\ell$ for a node $n$ into two shares: $\pi_1$ is held by $\ell$ itself, and $\pi_2$ is held by the lock of $n$'s parent. Then we amend definition \ref{inv1} to
\[R(n) \triangleq \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n')\]
Now the same sequence of operations gives us:
\[\begin{array}{l}
\{ \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n') \}\\
\texttt{acquire}(\ell');\\
\{ \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n') * \ell' \islock_{\pi_1} R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_{\pi_2} R(n'') \}\\
\texttt{release}(\ell);\\
\{ \ell' \islock_{\pi_1} R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_{\pi_2} R(n'') \}
\end{array}\]
When we acquire $\ell'$, we hold both $\pi_1$ and $\pi_2$ of its lock assertion; when we release $\ell$, we return share $\pi_1$ of $\ell$ along with share $\pi_2$ of $\ell'$, but retain share $\pi_1$ of $\ell'$, so we can release it later. The mechanism for defining a lock invariant that includes a share of its own lock assertion may vary across separation logics (VST provides a $\mathsf{selflock}$ assertion for exactly this purpose), but as long as such a mechanism exists, this pattern can be used to implement hand-over-hand lock invariants with share accounting, allowing us to traverse a data structure built with this pattern and then reclaim all of its resources once all threads are finished with it.

@@

\subsection{Specification of the hand-over-hand BST}
The C implementation of a node in our BST is:
\begin{lstlisting}[language = C,numbers = none]
typedef struct tree {int key; void *value;
                         struct tree_t *left, *right;} tree;
typedef struct tree_t {tree *t; lock_t *lock;} tree_t;
\end{lstlisting}
Each node (of type \lstinline{tree_t}) has a \lstinline{lock} field that holds the lock protecting the node, and a \lstinline{t} field that is either \lstinline{NULL} (for a leaf, which contains no key or value) or points to a \lstinline{tree} struct containing the node's key, value, and child pointers. To verify operations on this structure, we need to relate a \lstinline{tree_t} in memory to an abstract state (a mathematical map from keys to values), and divide ownership of those resources between the assertions $\nodeboxrep$ (the resources held by each client thread) and $\treerep$ (the abstract state of the data structure). As described in section \ref{atomicity}, our approach is to associate a piece of ghost state with each node in the tree, which will be shared between that node's lock and the abstract state in $\treerep$, while $\nodeboxrep$ will be a a pointer to the lock of the root node. We also need to make sure that our lock invariants support the hand-over-hand pattern.

\subsubsection{Key Ranges}%can cut this down if need be
When we traverse the BST looking for a key $k$, we will reach either a node containing $k$, or an empty node where $k$ would appear if it was in the tree. To prove correctness of this procedure, we need ghost state that tracks where each key ``should appear'' in the current tree. As observed by Krishna et al.~\cite{krishna2017flow}, this can be done by associating each node with a lower and upper bound on the keys appearing in the subtree rooted at that node. We refer to the product of these bounds as a \emph{range}. The range of a node is inherited from its parent, based on the parent's key: if node $n$ has range $(l, r)$ and key $k$, then its left child has range $(l, k)$ and its right child has range $(k, r)$. Figure~\ref{range_bst} shows an example of a BST with each node labeled with its range, starting with $(-\infty, +\infty)$ at the root and propagated to the empty leaf nodes.

@@

\begin{figure}[htb]
\centering
%% \forestset{
%%    lab/.style={
%%         label={[anchor=west, scale=0.75] north east:#1},
%%     },
%% }
\begin{forest}
for tree={
  calign=fixed edge angles,
  calign angle=65,
  circle, draw, minimum size=3ex, inner sep=1pt, l sep=0, l=0,
%%     s sep=3mm, l sep=0mm
        }
    % my label/.style={
        % label={[anchor=south west] north east:#1},
    % },
    [ 20, tikz={\node[right=0pt of .north east, scale=0.75]  {$(-\infty,\infty)$};}
        [15, tikz={\node[left=0pt of .north west, scale=0.75]  {$(-\infty,20)$};}]
        [30, tikz={\node[right=0pt of .north east, scale=0.75]  {$(20,\infty)$};}
            [25, tikz={\node[left=-2pt of .west, scale=0.75]  {$(20,30)$};}]
            [50, tikz={\node[right=0pt of .north east, scale=0.75]  {$(30,\infty)$};}
                [45, tikz={\node[right=0pt of .south east, scale=0.75]  {$(30,50)$};}
                    [40, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,45)$};}
                        [35, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,40)$};}
                            [, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,35)$};}]
                            [, tikz={\node[right=0pt of .north east, scale=0.75]  {$(35,40)$};}]
                        ]
                        [,no edge, draw=none]
                    ]
                    [,no edge, draw=none]
                ]
                [,no edge, draw=none]
            ]
        ]
    ]
\end{forest}
\caption{A BST with each node labeled with its range; empty children of nodes other than 35 are omitted}
\label{range_bst}
\end{figure}
